The Etiology of Cybersecurity

### 2022-03-18 reports/cimss_2022/main-cut.tex
cimss reduced to 18 pages

%\documentclass[conference]{llncs}
\documentclass[runningheads]{llncs}
%\usepackage{minipage}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[style=numeric]{biblatex}
\usepackage{amsmath}
\usepackage{amsfonts}
\addbibresource{../literature.bib}% Syntax for version >= 1.2
\usepackage{graphicx}
\graphicspath{{./images/}}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{tikz}
%\usepackage{amsthm}
\usepackage{rotating}

@@

\section{Introduction}\label{sec:intro}
\paragraph{Context and motivation.}
A \emph{scientific theory} is an explanation of a phenomenon such that the
explanation follows the scientific method. The \emph{scientific method} is an
\emph{empirical} method that aims at mitigating potential fallacies in
theories.  Karl Popper famously argued (e.g. in \autocite{Popper1959logic})
that a scientific theory can never be verified but only falsified, that a
theory should not be conceived by using the principle of
induction\footnote{Albert  Einstein wrote to Popper: ``[...] and I think (like
you, by the way) that theory cannot be fabricated out of the results of
observation, but that it can only be invented.'' \autocite{Popper1959logic}},
and that empirical experiments should be considered as the only evidence to
support the non-falseness of a scientific theory. 
In \autocite{Herley2016unfalsifiability}, Cormac Herley explores what he calls
``an asymmetry in computer security'', which he defines as follows: ``Things
can be declared insecure by observation, but not the reverse. There is no
observation that allows us to declare an arbitrary system or technique
secure''. With security, Herley only focuses on cybersecurity (we also use
security and insecurity, in this paper, only to refer to cyber-insecurity and
cybersecurity) and his intuition is that there is no scientific theory that can
predict the cybersecurity of a system, nor a theory that can predict all
possible insecurities of a system (which, by negation, may be used as a theory
of cybersecurity).  Herley then uses this argument to show that ``claims that
any measure is necessary for security are empirically unfalsifiable''.

\paragraph{Contributions.} 
The goal of this paper is to lay the foundations of a scientific cybersecurity
theory.  We consider the problem raised by Herley not confined to ``computer
security'' but rather we reason on any abstract system (so that our scientific
hypothesis\footnote{In the reminder of this paper, we will use the word
hypothesis to refer to ``scientific hypothesis'' as a proposed scientific
theory that has not gone through an extensive series of tests. Instead, we
use ``logical theory'' to refer to a set of formal logical axioms.} may
be tested in any sound implementation such as networks, mechanical, cyber, or
cyber-physical system, or even a single computer or a single device such as a
hard-drive).  
%There is also an apparent inconsistency in
%\autocite{Herley2016unfalsifiability} that we seek to clarify before following
%(as we agree) the scientific path drawn by Herley: cybersecurity is defined as
%a set of abstract properties in many formal approaches to the investigation of
%the cybersecurity of systems.  Let's, for the sake of simplicity, consider the case
%of a system implementing a cybersecurity protocol (e.g. an authentication protocol).
%The proof of cybersecurity obtained from the formal verification of the design of
%the protocol (e.g. with respect to a formalization of the authentication
%property) is indeed falsifiable against the cybersecurity properties verified.  For
%example, in the protocol verification community, cybersecurity is often defined
%as a formalization of the high-level properties confidentiality, integrity, and
%authentication. The problem in such approaches is not the definition of what
%cybersecurity is but the generality of the results, since they tackle a
%specific step\footnote{In fact, engineering models such as the V-model start
%with the requirements definition, and physical and functional architecture
%design, before the logic of protocols is even considered.} of the engineering
%process (of systems) without considering many aspects related to the design or
%implementation of a protocol that may generate cybersecurity vulnerabilities later
%in the development process.  An infamous example is the KRAK attack on the WPA2
%4-way handshake protocol where the author wrote \autocite{krak2020web}: ``our
%attacks do not violate the security properties proven in formal analysis of the
%4-way handshake'' but ``the problem is that the proofs do not model key
%installation.'' The theories underlying the verification of cybersecurity protocols
%are based on assumptions which non-evidently apply to a general cybersecurity
%theory; otherwise the assumptions would clearly state the necessary element
%that needs to be considered (e.g. modeled) in order to obtain a secure system.
%As an example, the so called Dolev-Yao attacker model\footnote{The Dolev-Yao
%attacker model can be considered as abstract model of an attacker that can be
%used for the analysis of protocol specifications.} \autocite{Dolev1983security}
%only applies to specific instances (often called scenarios) and abstraction of
%the protocol.  This, in turn, creates a false sense of security since it
%requires non-justifiable assumptions on the abstraction of the system under
%verification.
%More specifically, for the formal security verification of a system in which
%two or more agents are communicating, a formalized scenario needs to be
%defined by a modeler who chooses (among others): (i) a scope of the
%formalization (e.g.  excluding the server that distributes the public key is
%often done when verifying the security of authentication protocols), (ii) the
%number of sessions (even tough some approaches do reason on an infinite number
%of sessions such as \autocite{Escobar2007maudenpa}),\fixnote{Luca}{Nowadays, most
%tools work for infinite sessions: Maude-NPA, Tamarin, ProVerif, OFMC.} (iii)
%honesty/dishonesty of the peers (e.g.  in the ASLan++ language
%\autocite{Oheimb2010aslan++}),\fixnote{Luca}{Nowadays, most tools allow for easy
%switch of the resulting sessions} and (iv) the abstraction of the
%cryptographic primitives (e.g.  ProVerif vs CryptoVerif
%\autocite{Blanchet2017symbolic}).\fixnote{Luca}{This is the only claim that, more
%or less, still holds, even though more and more tools do cryptoanalysis. The
%good news is that, I believe, you do not need (i)-(iv) at all, but you can
%simply say that by the nature of modeling a specification, you are abstracting
%away certain things (e.g., cryptography, actual message content, PKI, etc.),
%and the vulnerabilities might be hidden there} While many tools and theories
%improved those issues, there is no agreement on which should be the definitive
%approach (if any).\fixnote{Luca}{I don't think that this is important. In many
%fields of research you don't have a single definitive approach and it is good
%that it is so. I would remove this sentence, but only focus on the next point,
%which is the only important one of this discussion of protocols} 
%More importantly, some of the choices made by the
%modelers/engineers using those formal approaches may completely change the
%results of the formal verification of the system (an interesting example on how
%difficult it is to even just compare the approaches is given in
%\autocite{Cremers2009comparing}). 
%\begin{figure}[t]
%	\centering
%	\includegraphics[width=.7\columnwidth]{protocol-example.pdf}
%	\caption{Abstraction of an ad-hoc esemplificative protocol execution}
%	\label{fig:protocol-example}
%\end{figure}
%As an illustrative example, in Fig.~\ref{fig:protocol-example}, under the
%perfect cryptography assumption (i.e. assuming that the cryptographic
%primitives have no cybersecurity flaw), and assuming that no violation to any
%cybersecurity property occurs after message I), the freedom of choosing the
%scope determines that the flaws related to the dishonest impersonification of
%the Server may or may not be considered in the verification process.  This
%choice has a tremendous impact on the focus and findings of the verification of
%the cybersecurity of the protocol.  While this may seem to turn upon minutiae
%and foreseeable, this highlights the false sense of security that may derive
%from a non-falsifiable theory of system cybersecurity.  In other words, 
%As
%Herley mentions in \autocite{Herley2016usenixvideo}, the problem is in the way
%we reason on cybersecurity, and we can see it in the way we reply to simple
%cybersecurity (foundamental) questions.  Let us consider the general question
%``what is cybersecurity?''; 
%%as in \autocite{hintikka2013inquiry}, we can
%%translate the wh-question into an equivalent existentially quantified formula.
%%But if the formula is $\exists P. P(s)\rightarrow Secure(s)$, 
%%(where we ask if
%if we detail the question as:
%there exists a predicate $P$ such that, given any arbitrary system $s$, when
%$P$ holds for $s$ then $s$ can be considered secure?
%%), 
%When the answer is (for
%example) ``if the system is confidential, the system is secure'' the reasoning %formula
%becomes tautological. In fact, $P$ and $Secure$ can be considered as
%semantically equivalent (cybersecurity is confidentiality and vice-versa).
Instead of starting from reasoning on what makes a system secure or insecure,
we reason on what causes insecurities. 
We focus on insecurities only caused by
the exploitation of cybersecurity attacks, and we assume that achieving
cybersecurity means preventing all those attacks from being exploitable or
exploited.  Our hypothesis is that \emph{cybersecurity attacks are only caused by the
presence of errors in the design or implementation of a system} (i.e.
cybersecurity weaknesses).
With our approach, a list of weaknesses emerges from the mathematical
formulation of a system in a framework called $\abftheory$-framework, and
predicts 4 main classes of weaknesses. Those classes of weaknesses are used to
calculate all the insecurity configurations of all the components of a system,
obtaining a precise estimation of all potential cybersecurity-related risks in
any given system. Our hypothesis can be falsified by means of experiments,
testing if all the predicted weaknesses are present in the system under test,
or testing if other (not predicted by our hypothesis) weaknesses are present.
In fact, if any cybersecurity weaknesses were to be found in a system and not
predicted by our hypothesis, the hypothesis could be declared incomplete.  If a
cybersecurity weakness would be predicted by our hypothesis but found to be impossible
to realize, our hypothesis could be declared as wrong.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.35\columnwidth]{causality.pdf}
	\caption{Etiology of cybersecurity}
	\label{fig:causality}
\end{figure}

As depicted in Fig.~\ref{fig:causality}, in order to define a theory on cybersecurity we may
say that the presence of vulnerabilities is a necessary condition in order to
cause an attack in the system.  Those vulnerabilities are, in turn, caused by
the presence of weaknesses in the system.  Weaknesses are errors in the design
or implementation of a system.  Therefore, a theory on cybersecurity should
first predict the errors in a system design.

\section{A Cybersecurity Hypothesis in the $\abftheory$-framework}\label{sec:hypothesis}
To address the problem raised by Herley, we define how to
distinguish between a secure and an insecure system in the following steps 
of the engineering process:
\begin{enumerate}
	\item \emph{System Specification}: the \emph{functional and physical requirements} are defined.
	\item \emph{Architecture Design}: the specification is structured into \emph{functional and physical architectures}.
	\item \emph{Cybersecurity Risk Assessment}: potential \emph{weaknesses} (errors) and
		\emph{cybersecurity requirements} are identified.	
    %\item \emph{Design Refinement}, where the physical and functional
	%	architectures are extended with \emph{mitigations} for the
	%	weaknesses. Mitigations at this step details the semantics of
	%	the functional blocks (behaviors), the data flow, and communication logic.
	%\item \emph{Verification of Security Requirements}, where the data flow, behaviors, and communication protocols are verified against the cybersecurity requirements.
\end{enumerate}
We changed the focus from the attacker as the source of insecurity to the
potential design errors of a system. We now define a framework for the
definition of a system that we use to identify weaknesses as potential design
errors.

@@

As in \autocite{Santaca2016abf}, we define agents as a meronomy (an
hierarchy of Part-Whole relations) over the constituents that we previously defined
(assertions, beliefs, and facts), based on a standard definition of mereology,
i.e. based on the definition of parthood relation between \emph{parts}.  Due to
the necessity of considering different relations between parts (as we'll show afterwards)
we extend the mereology to a
mereo-topology \autocite{Smith1996mereotopology,Varzi1994mereotopology,Rachavelpula2017mereotopology},
considering the relations in Tab.~\ref{tab:rcc358}.  For the sake of
readability, we use the term \emph{region} both to refer to a mereological part
and to a topological region.  Our aim is to create a meronomy (hierarchy of
part-whole relations) instead of the taxonomies (categorization based on
discrete sets) such as the one provided in \autocite{NIST2020NVD,MITRE2020CVE}
so that we don't need to rely on a scoring system (such as the CVSS) to assign
a quantitative evaluation of the cybersecurity of each entry. Instead, we want
a precise calculation of the number of insecure configurations of a system to
emerge from the mathematical formulation of our cybersecurity hypothesis.

A mereotopology, as defined e.g. in \autocite{Rachavelpula2017mereotopology},
is a mathematical structure where the basic relation between regions
is the reflexive and symmetric relation \emph{Connects With},
that we use to order a universe of agents
$\agentuniverse$ (see in Tab.~\ref{tab:rcc358}).  We use the Region Connection
Calculus (RCC), as defined in \autocite{bennettLogics,improvingRCC}, to provide
an axiomatization of the mereo-topological concepts. In its broader definition,
the RCC theory is composed by eight axioms, and is known as RCC8 \autocite{Grutter2008rcc}. Using
RCC5 instead of RCC8 prevents us from considering tangential connections
between spatial regions. However, tangential connections in RCC8 can be
considered as special cases of the more general spatial relations considered in
RCC5.
In Tab.~\ref{tab:rcc}, we summarize the axioms of the RCC (see, e.g., \autocite{Grutter2008rcc}).  We can now define a system
over the mereotopology using the RCC calculus, as follows, where $\rcc(X,Y)$ on
two generic regions $X,Y$ represents one of the possible RCC relations between
$X$ and $Y$. We note that all the RCC relations are symmetric with the exception
of those that have an explicit (related) inverse.

@@

As in \autocite{Santaca2016abf}, it follows that, defining
a system with a fixed number of regions, there exists
an upper-bound to the number of possible configuration of a system, defined by
the possible relations between the different regions.
For completeness, we report in the next paragraph 
the calculation done by the authors in \autocite{Santaca2016abf}.

\paragraph{Number of different configurations of a system.}
The general formula to calculate the number of different types of agents is
$r^{\binom{n}{k}}$, where $r$ is the number of relations with arity $k$,
between $n$ different regions. Therefore, $r^{\binom{n}{k}}$
expresses the number of permutation of $r$
relations over ${\binom{n}{k}}$ elements with repetitions, 
with ${\binom{n}{k}}$ being the number of
$k$-ary combinations of $n$ regions.
In our case, $\binom{n}{k}=3$ since we consider $3$ regions 
($\assertionRegion,\beliefRegion,\factRegion$), and all the relations
considered in the RCC are binary.  Hence, using RCC5 (with five different
spatial relations) over three regions, we can theoretically define up to 125
different type of agents. However, only 54 of the 125 (as showed in
\autocite{improvingRCC}) combinations are topologically correct with respect to
the definition of the relations of RCC5. Generalizing to all the RCCs:

\begin{itemize}%[nosep]
\item \emph{RCC3} --- theoretical: $3^3=27$,  correct: 15.
\item \emph{RCC5} --- theoretical: $5^3=125$, correct: 54.
%\item \emph{RCC8} --- theoretical: $8^3=512$, correct: 193.
\end{itemize}

%Hence, even if considering a different number of regions than the three
%$\assertionRegion$, $\beliefRegion$, and $\factRegion$ exponentially affects
%the number of theoretical agents, the application of RCC downscales that number
%of a factor that ranges from 1.8 to 2.5. In addition, using RCC5 we consider
%3.6 times more (different) types of agents than RCC3, but using RCC8 would
%allow us to consider 3.5 times more different agents.
In the quantitative evaluation of a single agent, depicted in Fig.~\ref{fig:quantitative},
we argue that only 1 configuration represents the nominal (expected) behavior 
of the agent while the other configurations are either impossible to 
implement or diverge from the intended nominal behavior. We note 
that the numbers reported here do not consider the details of the
engineering process and should be considered a limit of an abstract 
representation of the system.
\begin{figure}[t]
	\centering
	\includegraphics[width=.8\columnwidth]{quantitative.pdf}
	\caption{Cybersecurity risk for a single agent}
	\label{fig:quantitative}
\end{figure}

@@

\begin{adjustbox}{width=.7\columnwidth}

@@

\begin{figure}[t]
	\includegraphics[width=.5\columnwidth]{soundness.pdf}
	\caption{Example relation between facts, and assertions and beliefs}
	\label{fig:soundness}
\end{figure}

\paragraph{\Rcc{$\assertionRegion$}{$\beliefRegion$} -- Collaboration.} In order
to reason on the relation between assertions and beliefs we first need to
consider that, by definition, assertions are defined as transfer of information
between two agents.  Therefore, as depicted in
Fig.~\ref{fig:system-agent}, an agent has two main categories of assertions,
input and output assertions.  Given an agent $a$ and a collection of asserted
predicates $\Phi$,
the input assertions are those received by $a$ from an agent
$s$ acting as a sender, $\rassert{s}{a}{\Phi}$; similarly, output assertions
are sent from $a$ to a receiver $r$, $\rassert{a}{r}{\Phi}$. With a slight abuse
of notation, in the text we simply write $\assertionRegion_{s\rightarrow a}$ when we abstract
away from the content of the assertion (i.e. $\Phi$).
We shall consider
two pairs of regions:%\footnote{``I am not asserting, as Lotze did, that a
%relation between $X$ and $Y$ consists of a quality in $X$ and a quality in $Y$
%-- a view which I regard as quite indefensible. -- I assert that a relation $Z$
%between $X$ and $Y$ involves the existence in $X$ of the quality ``having the
%relation $Z$ to $Y$'' so that a difference of relations always involves a
%difference in quality, and a change of relations always involves a change of
%quality.'' --  Ellis J. McTaggart, The Unreality of
%Time \autocite{Mctaggart1908unreality}}
\begin{itemize}
	\item \Rcc{$\assertionRegion_{s\rightarrow a}$}{$\beliefRegion$},
		where the relation between input-assertions and beliefs
		describes the soundness of the execution of the functional
		architecture w.r.t. input elicitation.
		If all the inputs (assertions) are correctly
		handled in the functional specification (beliefs) the
		specification is complete.

@@

We note that our CSE define the properties of a secure system, and correlated
weaknesses can be found in the CWE dataset.  For example, CSE-$1$ can be seen as
correlated to the weakness class of ``Improper Interaction Between Multiple
Correctly-Behaving Entities'' defined by the CWE--$435$, CSE-$2$ with the
``Insufficient Control Flow Management'' defined by MITRE in the CWE--$691$,
and CSE-$3$ with the ``Improper Calculation'' defined by MITRE in the
CWE--$682$. All those CWE are in the top ``view'' of the ``research concepts''
in \autocite{MITRE2020CWEresearch}, while the other classes of weaknesses do not
have a direct counterpart in our hypothesis; we believe they can be sees as sub-classes,
but a full comparison with the CWE is out of the scope of this article.

We can now define what a secure system is (with respect to
the $\abftheory$-framework) and, based on that definition, what the cybersecurity risk is and
how to quantify it in a risk matrix. The following definition holds for 
abstract systems defined in the $\abftheory$-framework but will be refined
for CPS afterwards in the paper.
\begin{definition}{\bf Cybersecurity of a System or an Agent --}\label{def:security}
	A secure system is a system where CSE-$1$, CSE-$2$, and CSE-$3$ holds for
	each agent of the system.
\end{definition}
The ISO 31000 consider risk as the ``effect of uncertainty on objectives'' and
refers both to positive and negative consequences of uncertainty.
Accordingly, we consider risk as follows.
\begin{definition}{\bf Risk --}
The risk is the uncertainty related to the whole space of potential designs
	resulting from a specification in the $\abftheory$-framework.
\end{definition}
The definition of Risk leads to the risk matrix in
Fig.~\ref{fig:quantitative}, defined as follows.
\begin{definition}{\bf Risk Matrix --}
	The risk matrix, as summarized in Tab.~\ref{tab:5com}, is a function of the three relations
	$s=\langle\rcc(\factRegion,\beliefRegion),\rcc(\factRegion,\assertionRegion),\rcc(\beliefRegion,\assertionRegion)\rangle$,
	where the maximum risk is defined by the $DR$ relation between the
	three groups of regions, and the minimum risk by the $EQ$ relation over
	the same regions. In between the two extremes, the granularity of
	possible intermediate configuration is defined by the calculus used
	(RCC5 in our case).
\end{definition}
While a risk matrix is often defined as a function of the likelihood and impact
of attacks (based on quantitative ad-hoc estimation of how likely it is that an
attacker will exploit one or more vulnerabilities and what is the magnitude of
the incidents produced by this exploitation), we suggest that cybersecurity
weaknesses are all equally likely to be exploited if there's a connection
between the weakness and the asset that an attacker wants to impact. When the
whole system is considered an asset, all weaknesses are equally likely to be
exploited.  Therefore, a risk matrix should capture the number of insecure
configurations of a system, rather than predicating over likelihoods of
weaknesses exploitation.

@@

\subsection{From Multi-Agent to Cyber-Physical Systems}
\begin{figure}[t]
	\centering
	\includegraphics[width=.7\columnwidth]{abftheory.pdf}
	\caption{Mapping Epistemological Concepts to (Cyber-Physical) Systems Engineering}
	\label{fig:mashyp}
\end{figure}

>>>>

%\documentclass[conference]{llncs}
\documentclass[runningheads]{llncs}
%\usepackage{minipage}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[style=numeric-comp]{biblatex}
\usepackage{amsmath}
\usepackage{amsfonts}
\addbibresource{./literature.bib}% Syntax for version >= 1.2
\usepackage{graphicx}
\graphicspath{{./images/}}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{tikz}
%\usepackage{amsthm}
\usepackage{rotating}

@@

\section{Introduction}\label{sec:intro}
\paragraph{Context and motivation.}
A \emph{scientific theory} is an explanation of a phenomenon such that the
explanation follows the scientific method. The \emph{scientific method} is an
\emph{empirical} method that aims at mitigating potential fallacies in
theories.  Karl Popper famously argued (e.g. in \autocite{Popper1959logic})
that a scientific theory can never be verified but only falsified, that a
theory should not be conceived by using the principle of
induction\footnote{Einstein to Popper: ``[...] and I think (like
you, by the way) that theory cannot be fabricated out of the results of
observation, but that it can only be invented.'' \autocite{Popper1959logic}},
and that empirical experiments should be considered as the only evidence to
support the non-falseness of a scientific theory. 
In \autocite{Herley2016unfalsifiability}, Cormac Herley explores what he calls
``an asymmetry in computer security'', which he defines as follows: ``Things
can be declared insecure by observation, but not the reverse. There is no
observation that allows us to declare an arbitrary system or technique
secure''. With security, Herley only focuses on cybersecurity (we also use
security and insecurity, in this paper, only to refer to cyber-insecurity and
cybersecurity) and his intuition is that there is no scientific theory that can
predict the cybersecurity of a system, nor a theory that can predict all
possible insecurities of a system (which, by negation, may be used as a theory
of cybersecurity).  Herley then uses this argument to show that ``claims that
any measure is necessary for security are empirically unfalsifiable''.

\paragraph{Contributions.} 
The goal of this paper is to lay the foundations of a scientific cybersecurity
theory.  We consider the problem raised by Herley not confined to ``computer
security'' but rather we reason on any abstract system (so that our scientific
hypothesis\footnote{In the reminder of this paper, we will use the word
hypothesis to refer to ``scientific hypothesis'' as a proposed scientific
theory that has not gone through an extensive series of tests. We
use ``logical theory'' to refer to a set of formal logical axioms.} may
be tested in any sound implementation such as networks, mechanical, cyber, or
cyber-physical system, or even a single computer or a single device such as a
hard-drive).  
Instead of starting from reasoning on what makes a system secure or insecure,
we reason on what causes insecurities. 
We focus on insecurities only caused by
the exploitation of cybersecurity attacks, and we assume that achieving
cybersecurity means preventing all those attacks from being exploitable or
exploited.  Our hypothesis is that \emph{cybersecurity attacks are only caused by the
presence of errors in the design or implementation of a system} (i.e.
cybersecurity weaknesses).
With our approach, a list of weaknesses emerges from the mathematical
formulation of a system in a framework called $\abftheory$-framework, and
predicts 4 main classes of weaknesses. Those classes of weaknesses are used to
calculate all the insecurity configurations of all the components of a system,
obtaining a precise estimation of all potential cybersecurity-related risks in
any given system. Our hypothesis can be falsified by means of experiments,
testing if all the predicted weaknesses are present in the system under test,
or testing if other (not predicted by our hypothesis) weaknesses are present.
In fact, if any cybersecurity weaknesses were to be found in a system and not
predicted by our hypothesis, the hypothesis could be declared incomplete.  If a
cybersecurity weakness would be predicted by our hypothesis but found to be impossible
to realize, our hypothesis could be declared as wrong.

@@

\begin{figure}[t]
\begin{minipage}[t]{0.4\textwidth}
	\includegraphics[width=\columnwidth]{causality.pdf}
\end{minipage}
\begin{minipage}[t]{0.6\textwidth}
	\includegraphics[width=.9\columnwidth]{abftheory.pdf}
\end{minipage}
  \caption{Etiology of cybersecurity (left). Mapping Epistemological Concepts to (Cyber-Physical) Systems Engineering (right).}
	\label{fig:causality}
	\label{fig:mashyp}
\end{figure}

As in Fig.~\ref{fig:causality}, order to define a theory on cybersecurity we may
say that the presence of vulnerabilities is a necessary condition to
cause an attack in the system.  Those vulnerabilities are, in turn, caused by
the presence of weaknesses in the system.  Weaknesses are errors in the design
or implementation of a system and a theory on cybersecurity should
first predict the errors in a system design.

\section{A Cybersecurity Hypothesis in the $\abftheory$-framework}\label{sec:hypothesis}
To address the problem raised by Herley, we define how to
distinguish between a secure and an insecure system in the following steps 
of the engineering process:
\begin{enumerate}
	\item \emph{System Specification}: the \emph{functional and physical requirements} are defined.
	\item \emph{Architecture Design}: the specification is structured into \emph{functional and physical architectures}.
	\item \emph{Cybersecurity Risk Assessment}: potential \emph{weaknesses} (errors) and
		\emph{cybersecurity requirements} are identified.	
\end{enumerate}
We changed the focus from the attacker as the source of insecurity to the
potential design errors of a system. We now define a framework for the
definition of a system that we use to identify weaknesses as potential design
errors.

@@

As in \autocite{Santaca2016abf}, we define agents as a meronomy (an
hierarchy of Part-Whole relations) over the constituents that we previously defined
(assertions, beliefs, and facts), based on a standard definition of mereology,
i.e. based on the definition of parthood relation between \emph{parts}.  Due to
the necessity of considering different relations between parts (as we'll show afterwards)
we extend the mereology to a
mereo-topology \autocite{Smith1996mereotopology,Varzi1994mereotopology,Rachavelpula2017mereotopology},
considering the relations in Tab.~\ref{tab:rcc358}.  For the sake of
readability, we use the term \emph{region} both to refer to a mereological part
and to a topological region.  Our aim is to create a meronomy (hierarchy of
part-whole relations) instead of the taxonomies (categorization based on
discrete sets) such as the one provided in \autocite{NIST2020NVD,CVE}
so that we don't need to rely on a scoring system (such as the CVSS) to assign
a quantitative evaluation of the cybersecurity of each entry. Instead, we want
a precise calculation of the number of insecure configurations of a system to
emerge from the mathematical formulation of our cybersecurity hypothesis.
A mereotopology, as defined in \autocite{Rachavelpula2017mereotopology},
is a mathematical structure where the basic relation between regions
is the reflexive and symmetric relation \emph{Connects With},
that we use to order a universe of agents
$\agentuniverse$ (see in Tab.~\ref{tab:rcc358}).  We use the Region Connection
Calculus (RCC), as defined in \autocite{bennettLogics,improvingRCC}, to provide
an axiomatization of the mereo-topological concepts. In its broader definition,
the RCC theory is composed by eight axioms, and is known as RCC8 \autocite{Grutter2008rcc}. Using
RCC5 instead of RCC8 prevents us from considering tangential connections
between spatial regions. However, tangential connections in RCC8 can be
considered as special cases of the more general spatial relations considered in
RCC5.
In Tab.~\ref{tab:rcc}, we summarize the axioms of the RCC (see, e.g., \autocite{Grutter2008rcc}).  We can now define a system
over the mereotopology using the RCC calculus, as follows, where $\rcc(X,Y)$ on
two generic regions $X,Y$ represents one of the possible RCC relations between
$X$ and $Y$. We note that all the RCC relations are symmetric with the exception
of those that have an explicit (related) inverse.

@@

As in \autocite{Santaca2016abf}, it follows that, defining
a system with a fixed number of regions, there exists
an upper-bound to the number of possible configuration of a system, defined by
the possible relations between the different regions.
For completeness, we report in Appendix~\ref{app:agcalc}
some details on the calculation done by the authors in \autocite{Santaca2016abf}.
The general formula to calculate the number of different types of agents is
$r^{\binom{n}{k}}$, where $r$ is the number of relations with arity $k$,
between $n$ different regions. In our case, $\binom{n}{k}=3$ since we consider $3$ regions 
($\assertionRegion,\beliefRegion,\factRegion$), and all the relations
considered in the RCC are binary.  Hence, we have up to 125
different type of agents but only 54 of the 125 (as showed in
\autocite{improvingRCC}) combinations are topologically correct. 
Generalizing to all the RCCs:

\begin{itemize}%[nosep]
\item \emph{RCC3} --- theoretical: $3^3=27$,  correct: 15.
\item \emph{RCC5} --- theoretical: $5^3=125$, correct: 54.
\end{itemize}

%Hence, even if considering a different number of regions than the three
%$\assertionRegion$, $\beliefRegion$, and $\factRegion$ exponentially affects
%the number of theoretical agents, the application of RCC downscales that number
%of a factor that ranges from 1.8 to 2.5. In addition, using RCC5 we consider
%3.6 times more (different) types of agents than RCC3, but using RCC8 would
%allow us to consider 3.5 times more different agents.
In the quantitative evaluation of a single agent, as in Fig.~\ref{fig:quantitative},
we argue that only 1 configuration represents the nominal (expected) behavior 
of the agent while the other configurations are either impossible to 
implement or diverge from the intended nominal behavior. We note 
that the numbers reported here do not consider the details of the
engineering process and should be considered a limit of an abstract 
representation of the system.

\begin{figure}[t]
\begin{minipage}[t]{0.5\textwidth}
\includegraphics[width=.9\columnwidth]{quantitative.pdf}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
\includegraphics[width=.9\columnwidth]{soundness.pdf}
\end{minipage}
  \caption{Cybersecurity risk for a single agent (left). Example relation between facts, and assertions and beliefs.}
\label{fig:quantitative}
\label{fig:soundness}
\end{figure}

@@

\begin{adjustbox}{width=.6\columnwidth}

@@

\paragraph{\Rcc{$\assertionRegion$}{$\beliefRegion$} -- Collaboration.} 
By definition, assertions are defined as transfer of information
between two agents. An agent has two main categories of assertions,
input and output assertions.  Given an agent $a$ and a collection of asserted
predicates $\Phi$,
the input assertions are those received by $a$ from an agent
$s$ acting as a sender, $\rassert{s}{a}{\Phi}$; similarly, output assertions
are sent from $a$ to a receiver $r$, $\rassert{a}{r}{\Phi}$. With a slight abuse
of notation, in the text we simply write $\assertionRegion_{s\rightarrow a}$ when we abstract
away from the content of the assertion (i.e. $\Phi$).
We shall consider
two pairs of regions:
\begin{itemize}
	\item \Rcc{$\assertionRegion_{s\rightarrow a}$}{$\beliefRegion$},
		where the relation between input-assertions and beliefs
		describes the soundness of the execution of the functional
		architecture w.r.t. input elicitation.
		If all the inputs (assertions) are correctly
		handled in the functional specification (beliefs) the
		specification is complete.

@@

We note that our CSE define the properties of a secure system, and correlated
weaknesses can be found in the CWE dataset.  For example, CSE-$1$ can be seen as
correlated to the weakness class of ``Improper Interaction Between Multiple
Correctly-Behaving Entities'' defined by the CWE--$435$, CSE-$2$ with the
``Insufficient Control Flow Management'' defined by MITRE in the CWE--$691$,
and CSE-$3$ with the ``Improper Calculation'' defined by MITRE in the
CWE--$682$. All those CWE are in the top ``view'' of the ``research concepts''
in \autocite{MITRE2020CWEresearch}, while the other classes of weaknesses do not
have a direct counterpart in our hypothesis; we believe they can be sees as sub-classes,
but a full comparison with the CWE is out of the scope of this article.
We can now define what a secure system is (with respect to
the $\abftheory$-framework) and, based on that definition, what the cybersecurity risk is and
how to quantify it in a risk matrix. The following definition holds for 
abstract systems defined in the $\abftheory$-framework but will be refined
for CPS afterwards in the paper.
\begin{definition}{\bf Cybersecurity of a System or an Agent --}\label{def:security}
	A secure system is a system where CSE-$1$, CSE-$2$, and CSE-$3$ holds for
	each agent of the system.
\end{definition}
The ISO 31000 consider risk as the ``effect of uncertainty on objectives'' and
refers both to positive and negative consequences of uncertainty.
Accordingly, we consider risk as follows.
\begin{definition}{\bf Risk --}
The risk is the uncertainty related to the whole space of potential designs
	resulting from a specification in the $\abftheory$-framework.
\end{definition}
The definition of Risk leads to the risk matrix in
Fig.~\ref{fig:quantitative}, defined as follows.
\begin{definition}{\bf Risk Matrix --}
	The risk matrix, as summarized in Tab.~\ref{tab:5com}, is a function of the three relations
	$s=\langle\rcc(\factRegion,\beliefRegion),\rcc(\factRegion,\assertionRegion),\rcc(\beliefRegion,\assertionRegion)\rangle$,
	where the maximum risk is defined by the $DR$ relation between the
	three groups of regions, and the minimum risk by the $EQ$ relation over
	the same regions. In between the two extremes, the granularity of
	possible intermediate configuration is defined by the calculus used
	(RCC5 in our case).
\end{definition}
While a risk matrix is often defined as a function of the likelihood and impact
of attacks (based on quantitative ad-hoc estimation of how likely it is that an
attacker will exploit one or more vulnerabilities and what is the magnitude of
the incidents produced by this exploitation), we suggest that cybersecurity
weaknesses are all equally likely to be exploited if there's a connection
between the weakness and the asset that an attacker wants to impact. When the
whole system is considered an asset, all weaknesses are equally likely to be
exploited.  Therefore, a risk matrix should capture the number of insecure
configurations of a system, rather than predicating over likelihoods of
weaknesses exploitation.

@@

\subsection{From Multi-Agent to Cyber-Physical Systems}
As in Fig.~\ref{fig:mashyp}, we relate MAS and CPS as follows.
\begin{itemize}
	\item We consider a System as a hierarchy of agents. So, we map agents
		to \emph{systems, sub-systems, or devices}, depending on the
		granularity of the design. For example, a modeler can model a
		specific device as a system not decomposed into sub-systems or devices (and
		the device is considered as an agent).
	\item Agents reason over beliefs (i.e. transform beliefs into other
		beliefs) and each \emph{component of a CPS} (system, sub-system,
		or device) is composed by a \emph{functional architecture} that
		transforms input-beliefs into output-beliefs.
	\item Components of a CPS have \emph{ports} to exchange information
		with the outer environment (which may be a sub-system),
		similarly to agents. The transfer of information in a CPS is
		defined by \emph{channels}.
	\item The concept of facts is related to the \emph{requirements}
		that describe how the CPS shall behave and
		its \emph{physical architecture}.
